---
title: "Event Handling in PhysioExperiment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event Handling in PhysioExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Physiological experiments typically involve discrete events such as stimulus
presentations, participant responses, and experimental condition markers.
PhysioExperiment provides the `PhysioEvents` class and a set of functions for
creating, querying, filtering, and modifying event information within
`PhysioExperiment` objects.

This vignette covers the event system in detail: creating events, attaching
them to data objects, querying and filtering, and converting between time
and sample representations.

## The PhysioEvents Class

`PhysioEvents` is a lightweight S4 class that stores event data as a
`DataFrame` with four columns:

- **onset**: event onset time in seconds
- **duration**: event duration in seconds
- **type**: event category (e.g., "stimulus", "response")
- **value**: event label or value (e.g., "target", "hit")

### Creating Events

```{r create-events}
library(PhysioExperiment)

# Create events with full specification
events <- PhysioEvents(
  onset = c(1.0, 2.5, 4.0, 5.5, 7.0),
  duration = c(0.5, 0.5, 0.5, 0.5, 0.5),
  type = c("stimulus", "response", "stimulus", "response", "stimulus"),
  value = c("target", "hit", "distractor", "false_alarm", "target")
)
events
```

### Default Values

When creating events, only `onset` is strictly required. Other fields
receive sensible defaults:

```{r create-defaults}
# Duration defaults to 0, type to "event", value to ""
simple_events <- PhysioEvents(
  onset = c(1, 2, 3, 4, 5)
)
simple_events

# Single type is recycled across all events
stim_events <- PhysioEvents(
  onset = c(1, 2, 3, 4, 5),
  type = "stimulus"
)
stim_events
```

### Inspecting Events

```{r inspect-events}
# Number of events
nEvents(events)

# The show method prints a summary
events
```

## Attaching Events to PhysioExperiment

Events are stored in the `metadata` slot of a `PhysioExperiment` object.
Use `setEvents()` to attach them.

```{r attach-events}
# Create a PhysioExperiment
pe <- PhysioExperiment(
  assays = list(raw = matrix(rnorm(2500), nrow = 2500, ncol = 4)),
  colData = S4Vectors::DataFrame(
    label = c("Fz", "Cz", "Pz", "Oz"),
    type = rep("EEG", 4)
  ),
  samplingRate = 250
)

# Attach events
pe <- setEvents(pe, events)

# Count events attached to the object
nEvents(pe)
```

### Setting Events from a Data Frame

You can also pass a plain `data.frame` to `setEvents()`, which will be
automatically converted to a `PhysioEvents` object:

```{r events-from-df}
event_df <- data.frame(
  onset = c(0.5, 1.5, 2.5),
  duration = c(0.2, 0.2, 0.2),
  type = c("stimulus", "stimulus", "stimulus"),
  value = c("A", "B", "A")
)

pe <- setEvents(pe, event_df)
getEvents(pe)
```

## Retrieving Events

Use `getEvents()` to retrieve events from a `PhysioExperiment`. You can
optionally filter by event type.

```{r get-events}
# First, set up events with mixed types
pe <- setEvents(pe, PhysioEvents(
  onset = c(1.0, 1.8, 2.5, 3.2, 4.0, 4.7),
  type = c("stimulus", "response", "stimulus", "response", "stimulus", "response"),
  value = c("target", "hit", "distractor", "correct_reject", "target", "hit")
))

# Get all events
all_events <- getEvents(pe)
all_events

# Get only stimulus events
stim <- getEvents(pe, type = "stimulus")
stim

# Get only response events
resp <- getEvents(pe, type = "response")
nEvents(resp)
```

## Adding Events

Use `addEvents()` to append new events to existing ones. Events are
automatically sorted by onset time.

```{r add-events}
# Start with stimulus events
pe <- PhysioExperiment(
  assays = list(raw = matrix(rnorm(2500), nrow = 2500, ncol = 4)),
  samplingRate = 250
)

pe <- addEvents(pe, onset = c(1, 3, 5), type = "stimulus", value = "target")
nEvents(pe)  # 3

# Add response events
pe <- addEvents(pe,
  onset = c(1.5, 3.4),
  type = "response",
  value = c("hit", "hit")
)
nEvents(pe)  # 5

# Events are sorted by onset time
getEvents(pe)
```

## Removing Events

Use `removeEvents()` to remove events by type, by index, or remove all
events.

```{r remove-events}
# Remove all response events
pe_stim_only <- removeEvents(pe, type = "response")
nEvents(pe_stim_only)

# Remove specific events by index
pe_fewer <- removeEvents(pe, indices = c(1, 2))
nEvents(pe_fewer)

# Remove all events
pe_no_events <- removeEvents(pe)
nEvents(pe_no_events)  # 0
```

## Time and Sample Conversion

PhysioExperiment provides functions to convert between time in seconds and
sample indices, which is useful for aligning events with signal data.

```{r time-conversion}
pe <- PhysioExperiment(
  assays = list(raw = matrix(rnorm(2500), nrow = 2500, ncol = 4)),
  samplingRate = 250
)

# Convert times to sample indices
sample_idx <- timeToSamples(pe, c(0.0, 1.0, 2.0))
sample_idx  # 1, 251, 501

# Convert sample indices back to times
times <- samplesToTime(pe, c(1, 251, 501))
times  # 0.0, 1.0, 2.0
```

## Working with Event Windows

A common workflow is to extract data around events. You can combine
event retrieval with time-based subsetting:

```{r event-windows}
# Set up experiment with events
pe <- PhysioExperiment(
  assays = list(raw = matrix(rnorm(25000), nrow = 25000, ncol = 4)),
  colData = S4Vectors::DataFrame(label = c("Fz", "Cz", "Pz", "Oz")),
  samplingRate = 250
)

pe <- addEvents(pe,
  onset = c(10, 30, 50, 70),
  type = "stimulus",
  value = "target"
)

# Extract a window around the first event
events_df <- getEvents(pe)
first_onset <- events_df@events$onset[1]

# Extract 1 second before to 2 seconds after the event
pe_epoch <- extractWindow(pe, tmin = first_onset - 1, tmax = first_onset + 2)
duration(pe_epoch)  # approximately 3 seconds
```

## Events and Time Concatenation

When combining `PhysioExperiment` objects along the time axis using
`rbindPhysio()`, event onsets in the second object are automatically
offset by the duration of the first:

```{r rbind-events}
pe1 <- PhysioExperiment(
  assays = list(raw = matrix(rnorm(1000), nrow = 1000, ncol = 4)),
  samplingRate = 250
)
pe1 <- addEvents(pe1, onset = c(1, 2), type = "stimulus")

pe2 <- PhysioExperiment(
  assays = list(raw = matrix(rnorm(1000), nrow = 1000, ncol = 4)),
  samplingRate = 250
)
pe2 <- addEvents(pe2, onset = c(1, 2), type = "stimulus")

pe_combined <- rbindPhysio(pe1, pe2)

# Events from pe2 are offset by the duration of pe1 (4 seconds)
getEvents(pe_combined)
```

## Summary

The PhysioExperiment event system provides:

- **PhysioEvents()**: constructor for event objects with onset, duration,
  type, and value fields
- **setEvents() / getEvents()**: attach and retrieve events from
  PhysioExperiment objects
- **addEvents() / removeEvents()**: incrementally modify the event set
- **nEvents()**: count events
- **timeToSamples() / samplesToTime()**: convert between time and sample
  representations

These tools form the foundation for event-related analyses such as
epoching and ERP computation in downstream packages.

## Session Info

```{r session-info}
sessionInfo()
```
